# 2018 FIE Workshop {-}

```{r include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE)
opts_chunk$set(fig.width = 6, out.width = "70%", fig.align = "center", fig.asp = 0.618)
```

Workshop conducted at the 2018 Frontiers in Education Conference (2018-10-03) in San Jose, CA. 

The goal of the workshop is to share our data, methods, and metrics for intersectional research in student persistence. The workshop is designed for R beginners.

## Pre-workshop setup {-}

- [Getting started](https://midfieldr.github.io/workshops/getting-started.html) for pre-workshop software installation instructions.
- [Preface](https://midfieldr.github.io/workshops/index.html) for an introduction to MIDFIELD and the workshop facilitators. 
 


## Description {-}

The goal of the workshop is to make MIDFIELD more accessible to the FIE community via **midfieldr**. On completing the workshop, participants should be able to 

- Describe key variables in the MIDFIELD data 
- Select academic programs and populations to study
- Use R to compute and graph persistence metrics (e.g., graduation rate) 
- Explain key features of effective data displays 
- Continue using **midfieldr** to study additional persistence metrics

Participants should be sufficiently familiar with their operating systems to install software and navigate directories, but prior experience with R is not required. 


## Agenda  {-}

Workshop activities include think-pair share, active learning, demonstration, discussion, and self-paced software tutorials. Our 3-hour agenda includes: 

```{r}
suppressPackageStartupMessages(library(tidyverse))
df <- tribble(
  ~ Min, ~Topic, 
   10, "Introductions", 
   30, "Designing effective data displays (interactive)", 
   20," Getting started with R (interactive)", 
   20, "Break", 
   20, "Exploring the structure of the MIDFIELD data sample", 
   45, "Using R for persistence metrics (self-paced tutorial)", 
   15, "Additional persistence metrics using R", 
   20, "Discussing next steps and assessing the workshop"
  )
kable(df)
```



## Tutorial: Getting started with R  {-}

Open RStudio. 

In RStudio,

- File > New Project > New Directory > New Project 
- Fill in the text box Directory Name with 2018-FIE-R-Workshop
- Choose and directory on your computer to save the new project 
Select a location on your computer to save the project
Click Create Project to create a new directory



