# MIDFIELD Pre-Institute Workshop {#day-0} 

<!-- This file is included in the book only if it is listed in the _bookdown.yml file. -->

```{r include = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
opts_chunk$set(fig.width = 6, 
        out.width = "70%", 
        fig.align = "center", 
        fig.asp   = 0.618, 
        comment   = "#>", 
        warning   = FALSE, 
        message   = FALSE)
```

Agenda [[link](#day0agenda)]  

This is an **optional session** designed for R novices. If you cannot attend this session on Sunday, you are welcome to work these tutorials on your own before the Monday/Tuesday sessions. 

The tutorials give the R novice a quick introduction to three essential elements of data science using R:

- [R basics](#r-basics) 
- [Graph basics](#graph-basics) 
- [Data basics](#data-basics) 

The tutorials are designed to be completed by an R novice in less than 50 minutes each. The timing has been student-tested, but of course your mileage may vary. 







<br>
[&#9650; top of page](#day-0) 

```{r child = "R-basics.Rmd", echo = FALSE}
# imports the R basics Rmd script 
```

Feel free to take a break before starting the next tutorial. 

<br>
[&#9650; top of page](#day-0) 





```{r child = "graph-basics.Rmd", echo = FALSE}
# imports the graph basics Rmd script 
```

Feel free to take a break before starting the next tutorial. 

<br>
[&#9650; top of page](#day-0)    





```{r child = "data-basics.Rmd", echo = FALSE}
# imports the data basics Rmd script 
```

<br>
[&#9650; top of page](#day-0)  


  



## References {-}

<div id="refs"></div>
