---
title:     "Introduction to midfieldr"
subtitle:  "2019 MIDFIELD Institute"
author:    "Richard Layton"
date:      "June 2019"
output:
  binb::metropolis:
    includes:
      in_header: "../manage/slides_preamble.tex"
fontsize:   12pt
bibliography: ["../references447.bib"]
csl: "../manage/journal-of-glaciology.csl"
nocite: | 
  @Layton+Long+Ohland:2018:midfielddata, @Layton+Long+Ohland:2018:midfieldr, @MIDFIELD:2019, @Ohland+Zhang+Thorndyke+Anderson:2004
---

```{r include = FALSE}
library("knitr")
# opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               fig.asp = 1/2.5)
library("tidyverse")
library("forcats")
library("midfieldr")
library("bibtex")
```

## In education, cross-sectional designs are typical


\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 55mm 75mm 45mm 70mm, width = \linewidth, keepaspectratio, clip = true]{visuals/cross-section-research.pdf}
\end{adjustwidth}








## Longitudinal studies offer some advantages

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 55mm 75mm 45mm 70mm, width = \linewidth, keepaspectratio, clip = true]{visuals/longitudinal-research.pdf}
\end{adjustwidth}







## MIDFIELD is a database for longitudinal studies

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 5mm 0mm 0mm, width = \linewidth, keepaspectratio,  clip = true]{visuals/midfieldfinal.pdf}
\end{adjustwidth}

\begin{itemize}
  \item 2.2 M undergraduate students at 23 US institutions
  \item Whole-population data from registrars, 1988--2017
  \item For more information about MIDFIELD  \href{https://engineering.purdue.edu/MIDFIELD/}{(link)} 
\end{itemize}






## MIDFIELD data are curated in four categories

\vertskip{-3.8}

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 85mm 0mm 45mm, width = 0.96\linewidth, keepaspectratio, clip = true]{visuals/make-figures-11.pdf}
\end{adjustwidth}

\begin{tabular}{@{} m{30mm} m{100mm} @{}}
\raggedleft\raisebox{-1mm}{\includegraphics[height = 8mm]{visuals/logo-SAS.png}}
&
\LARGE\alertbf{2.2M students}
\end{tabular}





## R package midfielddata provides a stratified sample

\vertskip{-3.8}

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 85mm 5mm 45mm, width = 0.96\linewidth, keepaspectratio, clip = true]{visuals/make-figures-11b.pdf}
\end{adjustwidth}


\begin{tabular}{@{} m{30mm} m{100mm} @{}}
\raggedleft\raisebox{0mm}{\includegraphics[height = 10mm]{visuals/logo-R.png}}
&
\LARGE\alertbf{98,000 students}
\end{tabular}







## Each observation is \alert{a unique student}

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 30mm 5mm 45mm, width = 0.96\linewidth, keepaspectratio, clip=true]{visuals/make-figures-12.pdf}

\footnotesize\raggedright
\begin{tabular}{@{} p{0mm} p{1.2in} @{}}
& \texttt{\textbf{midfieldstudents}}\\
& 98,000 observations\\
& 19 Mb of memory
\end{tabular}
\end{adjustwidth}







## Each observation is \alert{one term for one student}

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 30mm 5mm 45mm, width = 0.96\linewidth, keepaspectratio, clip=true]{visuals/make-figures-13.pdf}

\footnotesize\raggedright
\begin{tabular}{@{} p{1.2in} p{1.1in}  @{}}
 & \texttt{\textbf{midfieldterms}}\\
 &   729,000 observations\\
 &   82 Mb of memory\\
\end{tabular}
\end{adjustwidth}












## Each observation is \alert{one course for one student}


\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 30mm 5mm 45mm, width = 0.96\linewidth, keepaspectratio, clip = true]{visuals/make-figures-14.pdf}

\footnotesize\raggedright
\begin{tabular}{@{} p{2.4in} p{1.1in}  @{}}
 & \texttt{\textbf{midfieldcourses}}\\
 &   3.5 M observations\\
 &   348 Mb of memory\\
\end{tabular}
\end{adjustwidth}






## Each observation is \alert{a unique student}

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 0mm 30mm 5mm 45mm, width = 0.96\linewidth, keepaspectratio, clip = true]{visuals/make-figures-15.pdf}

\footnotesize\raggedright
\begin{tabular}{@{} p{3.6in} p{1.2in}  @{}}
 & \texttt{\textbf{midfielddegrees}}\\
 &   98,000 observations\\
 &   10 Mb of memory
\end{tabular}
\end{adjustwidth}









## \alert{midfieldr} provides the functions for working with \alert{midfieldddata} 

\begin{adjustwidth}{-8mm}{-8mm}
\centering
\includegraphics[trim = 45mm 50mm 45mm 50mm, width = \linewidth, keepaspectratio, clip = true]{visuals/make-figures-02b-2.pdf}
\end{adjustwidth}







## For more information, see these package resources online 

\begin{adjustwidth}{-6mm}{-6mm}

\begin{tabular}{@{} p{7mm} p{200mm}  @{}}
\raggedleft\raisebox{-18mm}{\includegraphics[width = 14mm]{visuals/logo-midfieldr.png}}
 & 
\begin{itemize}
 \item  Package website \href{https://midfieldr.github.io/midfieldr/}{(link)} for more information and tutorials
 \item  Package reference page \href{https://midfieldr.github.io/midfieldr/reference/}{(link)} for descriptions of functions
 \item  Package source code \href{https://github.com/MIDFIELDR/midfieldr}{(link)} 
\end{itemize}\\

\raggedleft\raisebox{-18mm}{\includegraphics[width = 14mm]{visuals/logo-midfielddata.png}}
 & 
\begin{itemize}
 \item  Package website \href{https://midfieldr.github.io/midfielddata/}{(link)} for more information   
 \item  Package reference page \href{https://midfieldr.github.io/midfielddata/reference/}{(link)} for data dictionaries  
  \item  Package source code \href{https://github.com/MIDFIELDR/midfielddata}{(link)} 
\end{itemize}

\end{tabular}

\end{adjustwidth}













## References
\scriptsize


```{r}
# pdf slide byproducts
unlink("slides/*.tex")
unlink("slides/*.aux")
unlink("slides/*.snm")
unlink("slides/*.toc")
unlink("slides/*.out")
unlink("slides/*.nav")
unlink("slides/*.log")
```


