---
title:     "Elements of effective graphs"
subtitle:  "2019 MIDFIELD Institute"
author:    "Richard Layton"
date:      "June 2019"
output:
  binb::metropolis:
    includes:
      in_header: "../manage/slides_preamble.tex"
fontsize:   12pt
bibliography: ["../references447.bib"]
biblio-style: abbrvnat
nocite: | 
  @Tufte:1983, @Wainer:2000, @Robbins:2013, @Kostelnick+Hassett:2003
---

```{r include = FALSE}
library("knitr")
# opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               fig.asp = 1/2.5)
```


```{r echo = FALSE}
library("tidyverse")
library("forcats")
library("midfieldr")
library("bibtex")

# theme for these slides 
font_size  <- 24
font_color <- rcb("dark_Gray")
theme_slides <- theme_bw() + 
        theme(panel.grid.minor = element_blank(),  
              panel.grid.major = element_line(size = 1), 
                text = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.title = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.subtitle = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.caption  = element_text(size = font_size, 
                        face = "plain", color = font_color),
                axis.title    = element_text(size = font_size, 
                        face = "plain", color = font_color),
                axis.text     = element_text(size = font_size, 
                        face = "plain", color = font_color),
                legend.title  = element_text(size = font_size, 
                        face = "plain", color = font_color),
                legend.text   = element_text(size = font_size, 
                        face = "plain", color = font_color),
                strip.text    = element_text(size = font_size, 
                        face = "plain", color = font_color, hjust = 0)
        )
```





## In your handout, list the slices A thru E from largest to smallest


\begin{columns}[onlytextwidth]
  
  \column{0.6\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/slides-pie-1.pdf}
  
  \column{0.4\textwidth}
  
\begin{itemize}\Large
  \item<2->  \alert {\bf B} (largest)
  \item<3->  \alert {\bf D}
  \item<3->  \alert {\bf A}
  \item<3->  \alert {\bf C} 
  \item<3->  \alert {\bf E} (smallest)
\end{itemize}
  
\end{columns}

\vspace{2\baselineskip}
\scriptsize Adapted from [@Robbins:2013, Ch 2]







## The same data arranged along a common axis  

Comparing values along a common axis is a high-accuracy visual task. 
  
```{r}
df2 <- tribble(~N, ~pie, 
                23, "B", 
                21, "D", 
                20, "A", 
                19, "C", 
                17, "E") %>% 
        mutate(pie = fct_reorder(pie, N)) 

ggplot(df2, aes(x = N, y = pie)) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        labs(x = "", y = "") +
        scale_x_continuous(breaks = c(17, 18, 19, 20, 21, 22, 23)) +
        theme_slides
```





  
## Structured data. What patterns do you see in these data? 


\bigskip

\includegraphics[width = 0.9\linewidth]{visuals/structured-data.png} 


\vspace{0\baselineskip}
\scriptsize Adapted from [@Robbins:2013, Ch 2] 







## Structured data on a common scale. Any new observations? 

```{r fig.asp = 1/2}
df <- wrapr::build_frame(
        "label"     ,"item" , "value" |
        "Item 09"   ,   9   , 6.5     |
        "Item 01"   ,   1   , 7       |
        "Item 05"   ,   5   , 7.5     |
        "Item 03"   ,   3   , 8       |
        "Item 07"   ,   7   , 8.5     |
        "Item 10"   ,  10   , 11.5    |
        "Item 02"   ,   2   , 12      |
        "Item 06"   ,   6   , 12.5    |   
        "Item 04"   ,   4   , 13      |
        "Item 08"   ,   8   , 13.5    )

df <- df %>% 
  mutate(oddeven = ifelse(item %% 2 == 0, "even", "odd")) %>% 
  arrange(oddeven, value) %>% 
  mutate(cycle = rep(1:5, 2))


ggplot(data = df, mapping = aes(x = value, y = label)) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        labs(x = "", y = "") + 
        scale_x_continuous(limits = c(0, 15)) +
        theme_slides 
```






## Reorder the rows by the data values. New observations? 

```{r fig.asp = 1/2}
ggplot(data = df, mapping = aes(x = value, y = reorder(label, value))) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        labs(x = "", y = "") + 
        scale_x_continuous(limits = c(0, 15)) +
        theme_slides
```









## Suppose we move the item number to the data marker. 

```{r fig.asp = 1/2}
ggplot(data = df, mapping = aes(x = value, y = reorder(label, value))) +
        geom_point(size = 11, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 11, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        geom_text(mapping = aes(x = value, y = reorder(label, value), label = item), size = 7.5) +
        labs(x = "", y = "") + 
        scale_x_continuous(limits = c(0, 15)) +
        theme_slides 
```






## Even-odd pairs emerge 

```{r fig.asp = 1/3}
ggplot(data = df, mapping = aes(x = value, y = cycle)) +
        geom_point(size = 11, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 11, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        geom_text(mapping = aes(x = value, y = cycle, label = item), 
                  size = 7.5) +
        labs(x = "", y = "") + 
        scale_x_continuous(limits = c(0, 15)) +
        scale_y_continuous(limits = c(0.5, 5.5)) +
        theme_slides +
        theme(axis.text.y = element_blank())
```


\bigskip 
The greatest value of data visualization is when       
\alertbf{it forces us to notice what we never expected to see}.    
\raggedleft --- John Tukey (1915--2000)



  
 
## Slices are what percentage of the whole? 

\bigskip

\includegraphics[width=1\linewidth]{visuals/slides-3d-pie-1.pdf} 

  \begin{tabular}{@{} lr @{}}
    Fill in the blanks         & A. \Large\rule{0.5in}{0.5pt}\\
    The total should be 100\%  & B. \Large\rule{0.5in}{0.5pt}\\
                               & C. \Large\rule{0.5in}{0.5pt}\\
                               & D. \Large\rule{0.5in}{0.5pt}\\
  \end{tabular} 
  
\vspace{1\baselineskip}
\scriptsize Adapted from [@Robbins:2013, Ch 2]  
  
  
  
  
  

  
  
  
  
  
  
  
  
  
## 3D-effects distort our judgment  

\bigskip

\includegraphics[width=1\linewidth]{visuals/slides-3d-pie-1.pdf} 

  \begin{tabular}{@{} lr @{}}
    Fill in the blanks         & A. \quad \alert {\bf\Large 20\%}\\
    The total should be 100\%  & B. \quad \alert {\bf\Large 20\%}\\
                               & C. \quad \alert {\bf\Large 20\%}\\
                               & D. \quad \alert {\bf\Large 40\%}\\
  \end{tabular}  
  
\vspace{1\baselineskip}
\scriptsize Adapted from [@Robbins:2013, Ch 2] 






## Again, the same data arranged along a common axis  

A high-accuracy visual task. 
  
```{r}
df2 <- tribble(~N, ~pie, 
                20, "A", 
                20, "B", 
                20, "C", 
                40, "D") %>% 
        mutate(pie = fct_reorder(pie, N)) 

ggplot(df2, aes(x = N, y = pie)) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        labs(x = "", y = "") +
        scale_x_continuous(limits = c(18, 42)) +
        theme_slides
```
    
  
## Write down the heights of the bars 

\bigskip

\begin{columns}[onlytextwidth]
  
  \column{0.65\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/slides-3d-bar-1.pdf} 
  
  
  \column{0.05\textwidth}
  
  \column{0.3\textwidth}
  This is a visual inspection only. 
  
  \bigskip
  
  \begin{tabular}{@{} l @{}}
  Fill in the blanks\\  
  A. \Large\rule{0.5in}{0.5pt}\\
  B. \Large\rule{0.5in}{0.5pt}\\
  C. \Large\rule{0.5in}{0.5pt}\\
  D. \Large\rule{0.5in}{0.5pt}\\
  \end{tabular} 
  
\end{columns}

\vspace{0\baselineskip}
\scriptsize Adapted from [@Robbins:2013, p. 22]





##  Again, 3D-effects distort our judgment 

\bigskip

\begin{columns}[onlytextwidth]
  
  \column{0.65\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/slides-3d-bar-1.pdf} 
  
  
  \column{0.05\textwidth}
  
  \column{0.3\textwidth}
  This is a visual inspection only. 
  
  \bigskip
  
  \begin{tabular}{@{} l @{}}
  Fill in the blanks\\  
  A. \quad \alert {\bf\Large 2}\\
  B. \quad \alert {\bf\Large 4}\\
  C. \quad \alert {\bf\Large 6}\\
  D. \quad \alert {\bf\Large 8}\\
  \end{tabular} 
  
\end{columns}
  
  
  
  


<!-- ## Again, the same data arranged along a common axis   -->

<!-- A high-accuracy visual task.  -->

```{r}
df2 <- tribble(~N, ~pie,
                2, "A",
                4, "B",
                6, "C",
                8, "D") %>%
        mutate(pie = fct_reorder(pie, N))

# ggplot(df2, aes(x = N, y = pie)) +
#         geom_point(size = 7, alpha = 0.7, color = rcb("light_BG")) +
#         geom_point(size = 7, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
#         labs(x = "", y = "") +
#         # scale_x_continuous(limits = c(18, 42)) +
#         theme_slides
```

    
    
    
  
## You can use bars, but must include zero

```{r}
bar1 <- ggplot(df2, aes(x = pie, y = N)) +
        geom_col(color = rcb("mid_BG"), fill = rcb("mid_BG"), width = 0.35) +
        labs(x = "", y = "") +
        coord_flip() + 
        theme_slides + 
        theme(rect = element_rect(fill = "transparent"))
bar1
```







##  If you mark the endpoints, you can omit the bar

```{r}
bar2 <- bar1 +
        geom_col(color = rcb("light_BG"), fill = rcb("pale_BG"), width = 0.4) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG"))

bar2
```

  
  
  
  
  
  
 
##  Producing a "dot plot" with rows ordered per the data 

```{r}
ggplot(df2, aes(x = N, y = pie)) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        scale_x_continuous(limits = c(0, 8)) +
        labs(x = "", y = "") +
        theme_slides
```


  
  
  
  
  
  
  
## 1st attempt: Visually estimate the state areas  

\bigskip

Visual estimation of area is a low-accuracy task. 

\bigskip

\begin{columns}[onlytextwidth]
  
  \column{0.45\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/slides-states-1.pdf} 
  
  
  \column{0.05\textwidth}
  
  \column{0.55\textwidth}
South Carolina (SC) $\approx$ 83,000 sq km. 
  
  \bigskip
  
  \begin{tabular}{@{} l r l @{}}
  FL & \Large\rule{0.5in}{0.5pt} & \normalsize x 1000 sq. km\\
  GA & \Large\rule{0.5in}{0.5pt} &\normalsize x 1000 sq. km\\
  AL & \Large\rule{0.5in}{0.5pt} &\normalsize x 1000 sq. km\\
  SC & 83 &\normalsize x 1000 sq. km\\
  \end{tabular}
  
\end{columns} 
  
  
  
\vspace{0\baselineskip}
\scriptsize Adapted from [@Ihaka:2007] 
  
  
  
  
  
  
  
## 2nd attempt: Visually estimate the state areas  

  
```{r}
df3 <- tribble(~area, ~state, 
                82933, "SC", 
                170312, "FL", 
                153910, "GA", 
                135767, "AL") %>% 
        mutate(state = fct_reorder(state, area)) 

fig3 <- ggplot(df3, aes(x = area/1000, y = state)) +
        geom_point(size = 8, alpha = 0.7, color = rcb("light_BG")) +
        geom_point(size = 8, shape = 1, stroke = 1.5, color = rcb("dark_BG")) +
        labs(x = "", y = "") +
        scale_x_continuous(limits = c(80, 180), 
                breaks = c(80, 100, 120, 140, 160, 180)) +
        theme_slides + 
        theme(rect = element_rect(fill = "transparent"))

ggsave(filename = "slides-my-states.png", 
        path   = "visuals",
        width  = 10, 
        height = 10/3, 
        units  = "in", # "in", "cm", or "mm"
        dpi    = 600)
```

 
\bigskip

\includegraphics[width=1\linewidth]{visuals/slides-my-states.png} 

\begin{center}
  \begin{tabular}{@{} l r l @{}}
  FL & \Large\rule{0.5in}{0.5pt} & \normalsize x 1000 sq. km\\
  GA & \Large\rule{0.5in}{0.5pt} & \normalsize x 1000 sq. km\\
  AL & \Large\rule{0.5in}{0.5pt} & \normalsize x 1000 sq. km\\
  SC & 83 & \normalsize x 1000 sq. km\\
  \end{tabular}
\end{center}
  







  
## Your estimates have probably improved  

\bigskip

\includegraphics[width=1\linewidth]{visuals/slides-my-states.png} 

\begin{center}
  \begin{tabular}{@{} l r @{}}
  FL & \alert {\bf\Large 170} \normalsize x 1000 sq. km\\
  GA & \alert {\bf\Large 154} \normalsize x 1000 sq. km\\
  AL & \alert {\bf\Large 136} \normalsize x 1000 sq. km\\
  SC & \alert {\bf\Large 83} \normalsize x 1000 sq. km\\
  \end{tabular}
\end{center}







## Color represents \alert{geographic area}. What is the visual story?  

\begin{adjustwidth}{-2.9em}{0em}
\vertskip{13.3} 
\includegraphics[scale = 0.595, trim = 0mm 0mm 0mm 4mm, clip = TRUE]{../figures/election-map-2016-county-area.png}
\end{adjustwidth}
\FloatBarrier








## \alert{One dot} is \alert{one vote.} Color by party. What is the visual story?  

\begin{adjustwidth}{-2.9em}{0em}
\vertskip{13.3} 
\includegraphics[scale = 0.4, trim = 0mm 0mm 0mm 3mm, clip = TRUE]{../figures/election-map-2016-one-dot-one-vote.png}
\end{adjustwidth}
\FloatBarrier









## Which story does the President favor? 

\begin{adjustwidth}{-2.9em}{0em}
\vertskip{13.3} 
\includegraphics[scale = 0.43, trim = 0mm 0mm 0mm 0mm, clip = TRUE]{../figures/election-map-2016-oval-office.png}
\end{adjustwidth}
\FloatBarrier







## Area is \alert{county GDP}. Color by party. What is the visual story?

\begin{adjustwidth}{-2.9em}{0em}
\vertskip{13.2} 
\includegraphics[scale = 0.41, trim = 0mm 2mm 0mm 1mm, clip = TRUE]{../figures/election-map-2016-county-gdp.png}
\end{adjustwidth}
\FloatBarrier








## The numbers that \alert{decide the election}. What is the visual story?

\begin{adjustwidth}{-2.9em}{0em}
\vertskip{12} 
\includegraphics[scale = 0.41, trim = 0mm 0mm 0mm 0mm, clip = TRUE]{../figures/election-map-2016-electoral-votes.png}
\end{adjustwidth}
\FloatBarrier








## The previous graphs \alert{conceal what story} ?

```{r fig.asp = 1/2.25}
df <- wrapr::build_frame(
                "voter", "pct" |
                "Did not vote", 40  |
                "Voted for Trump", 28 |
                "Voted for Clinton", 29 |
                "Voted for other candidates", 3) %>% 
        mutate(voter = fct_reorder(voter, pct)) 






my_color   <- c(rcb("dark_BG"), "#BF0A30", "#002868", rcb("mid_Gray"))


ggplot(data = df, mapping = aes(x = pct, y = voter, color = voter)) +
        geom_point(size = 8, alpha = 0.7) +
        geom_point(size = 8, alpha = 1, shape = 1, stroke = 1.5) +
        scale_x_continuous(limits = c(0, 45)) +
        scale_color_manual(values = my_color) +
        labs(x = "\nPercent of eligible voters", y = "") +
        theme_slides +
        theme(legend.position = "none")
```















## The trouble with bars. What's wrong with this graph? 


\begin{columns}[onlytextwidth]
  
  \column{0.6\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/science-bars-1.png}
  
  \column{0.45\textwidth}
  

\end{columns}

\vspace{1\baselineskip}
\scriptsize  Norman L. Fortenberry, Jacquelyn F. Sullivan, Peter N. Jordan, and Daniel W. Knight (2007), Engineering education research aids instruction, *Science*, 31:1175--1176. 









## A nearly constant difference seems to increase in significance


\begin{columns}[onlytextwidth]
  
  \column{0.6\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/science-bars-4.png}
  
  \column{0.45\textwidth}
  
\begin{itemize}
  \item<2-> \alertbf{Can you identify the missing information?}
\end{itemize}
  
\end{columns}

\vspace{1\baselineskip}
\scriptsize  Norman L. Fortenberry, Jacquelyn F. Sullivan, Peter N. Jordan, and Daniel W. Knight (2007), Engineering education research aids instruction, *Science*, 31:1175--1176. 











## Redesigned, with full scales, a different story emerges

\bigskip

\begin{columns}[onlytextwidth]
  
  \column{0.6\textwidth}
  
  \bigskip
  
  \includegraphics[width=1\linewidth]{visuals/science-bars-3.png}
  
  \column{0.45\textwidth}
  
\begin{itemize}
  \item<2-> FYEP impact is in the first year. 
  \item<2-> Attrition rate afterwards is about the same for both groups.
\end{itemize}
  
\end{columns}









## The experts tell us 

\begin{columns}
\column{0.4\linewidth}
\includegraphics[width=1\linewidth]{visuals/doumont.jpg} 

\column{0.6\linewidth}
Optimal design primarily depends on    

\begin{itemize}
\item The message to be conveyed    
\item The variables to be shown   
\end{itemize}

\end{columns}

      
[@Doumont:2009]


\tiny  Image from http://www.principiae.be/pdfs/Principiae-2014.pdf









## The  experts tell us 

\begin{columns}
\column{0.4\linewidth}
\includegraphics[width=1\linewidth]{visuals/tufte.jpg} 

\column{0.6\linewidth}


The task of the designer is to give visual access to the subtle and the difficult --- that is, reveal the complex.




\end{columns}

\footnotesize      
[@Tufte:1983]

\tiny  Image from https://en.wikipedia.org/wiki/Edward_Tufte








## The  experts tell us 

\begin{columns}
\column{0.4\linewidth}
\includegraphics[width=1\linewidth]{visuals/evergreen.jpg} 

\column{0.6\linewidth}
What’s your point? 

$\qquad$

Seriously, that's the most important question. 


\end{columns}

\footnotesize 
[@Evergreen:2017]

\tiny  Image from https://tei.cgu.edu/people/stephanie-evergreen-phd/











## References
\scriptsize



