---
title:     "Expanding your graphical repertoire"
subtitle:  "2019 MIDFIELD Institute"
author:    "Richard Layton"
date:      "June 2019"
output:
  binb::metropolis:
    includes:
      in_header: "../manage/slides_preamble.tex"
fontsize:   12pt
bibliography: ["../references447.bib"]
biblio-style: abbrvnat
nocite: | 
---

```{r include = FALSE}
library("knitr")
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               fig.asp = 1/2.5)
```

```{r echo = FALSE}
library("tidyverse")
library("forcats")
library("midfieldr")
library("bibtex")

# theme for these slides 
font_size  <- 20
font_color <- rcb("dark_Gray")
theme_slides <- theme_bw() + 
        theme(panel.grid.minor = element_blank(),  
              panel.grid.major = element_line(size = 1), 
                text = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.title = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.subtitle = element_text(size = font_size, 
                        face = "plain", color = font_color),
                plot.caption  = element_text(size = font_size, 
                        face = "plain", color = font_color),
                axis.title    = element_text(size = font_size, 
                        face = "plain", color = font_color),
                axis.text     = element_text(size = font_size, 
                        face = "plain", color = font_color),
                legend.title  = element_text(size = font_size, 
                        face = "plain", color = font_color),
                legend.text   = element_text(size = font_size, 
                        face = "plain", color = font_color),
                strip.text    = element_text(size = font_size, 
                        face = "plain", color = font_color, hjust = 0)
        )
```










## \alert{Data structure} determines which graph types are feasible

\bigskip
\begin{adjustwidth}{-1.7em}{-1.7em}
\begin{tabular}{@{} >{\raggedright\arraybackslash} m{39mm} m{30mm} m{37mm} @{}}
&
\includegraphics[width = 30mm]{visuals/sd011-14.png}  
& 
Number of variables?\newline
Continuous or discrete?\newline
\quad\\

Number of observations?&&\\

&
\includegraphics[width = 30mm]{visuals/sd011-14b.png}  
& 
Number of variables?\newline
Nominal or ordinal?\newline
Number of levels each?
\end{tabular}
\end{adjustwidth}
\FloatBarrier





## \alert{Graphical repertoire} determines which graph types you explore

\begin{adjustwidth}{-3em}{0em}
\vertskip{-1.85} 
\includegraphics[scale=0.776]{visuals/sd011-13.png}
\end{adjustwidth}
\FloatBarrier

\vertskip{-13}
\begin{columns}[onlytextwidth]

   \begin{column}{0.58\textwidth}\bf\large
   strip plot\\
   box and whisker plot\\
   multiway\\
   scatterplot\\
   dot plot\\
   line graph\\
   conditioning plot\\
   scatterplot matrix\\
   parallel coordinate plot
   \end{column}

   \begin{column}{0.45\textwidth}\bf\large
   cycle plot\\
   quantile-quantile plot\\
   Sankey diagram\\
   diverging stacked bar\\
   multivariate bar\\
   micromaps\\
   proportional symbol map\\
   dot density map\\
   \quad\\
   \end{column}

\end{columns}













# Gallery of graph types 

 







## Strip plot --- or jitter plot or 1D scatterplot

```{r eval = TRUE}
speed_ski <- readRDS("data/0302-strip-plot-speedski-data.rds")
```

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Speed\newline    
Categorical: Event, sex
\end{adjustwidth}

\begin{adjustwidth}{-3em}{0em}
\vertskip{-0.5}
\includegraphics[width = 1.15\textwidth]{visuals/0302-strip-plot-speedski-slides.png}
\end{adjustwidth}

\scriptsize\raggedright
\vertskip{0.5}
Data from Unwin, Antony. 2015. GDAdata: Datasets for the book Graphical Data Analysis with R.









## Box and whisker plot or box plot 

```{r eval = TRUE}
nontraditional <- readRDS("data/0305-boxplot-nontrad-data.rds")
```

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Years enrolled\newline    
Categorical: Path + sex
\end{adjustwidth}

\begin{adjustwidth}{-3em}{0em}
\vertskip{-0.5}
\includegraphics[width = 1.15\textwidth]{visuals/0305-boxplot-nontrad-slides.png}
\end{adjustwidth}










## Multiway plot 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Graduation rate\newline   
Categorical: Race/ethnicity/sex, major 

\centering\includegraphics[height = 0.75\textheight]{visuals/fig004-grad-rate.png}
\end{adjustwidth}










## Scatterplot 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Migrators attracted, migrators graduated\newline    
Categorical: Race/ethnicity, sex, major 

\includegraphics[height = 0.7\textheight]{visuals/fig010-migr-yield.png}
\end{adjustwidth}
  










## Cleveland dot plot

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Stickiness\newline  
Categorical: Race/ethnicity/sex, major, graduation destination\newline 

\includegraphics[height = 0.45\textheight]{visuals/dot-plot-example.png}

\end{adjustwidth}
 






## Line graph 

\begin{adjustwidth}{-1.7em}{-1.7em}

\vertskip{0}
\includegraphics[height = 0.95\textheight]{visuals/d4-pollutionMortality-1.png}

\scriptsize\raggedright
\vertskip{-2} 
Thomas Ryan, 2019 portfolio    

\end{adjustwidth}






 









## Conditioning plot or co-plot 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Ozone, temperature, wind speed, solar radiation  

\centering
\includegraphics[height = 0.85\textheight]{visuals/co-plot.png}

\end{adjustwidth}








## Scatterplot matrix 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Six dimensions of Swiss banknotes 

\centering
\includegraphics[height = 0.85\textheight]{visuals/scatterplot-matrix.png}

\end{adjustwidth}






## Parallel coordinate plot

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Six dimensions of Swiss banknotes \newline

\centering
\includegraphics[height = 0.65\textheight, trim = 0mm 0mm 0mm 7mm, clip = TRUE]{visuals/parallel-coord.png}

\end{adjustwidth}






## Cycle plot or month plot 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Number of wildlife strikes on aircraft \newline
Categorical: Month, year 

\centering
\includegraphics[height = 0.68\textheight, trim = 0mm 0mm 0mm 22mm, clip = TRUE]{visuals/cycle-plot.png}

\raggedright\scriptsize
\vertskip{-1}
https://canonicalized.com/cycle-plots-tableau/
\end{adjustwidth}








## Quantile-quantile plot or q-q plot 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Years to graduate quantiles\newline   
Categorical: Traditional students, nontraditional students 

\centering\includegraphics[height = 0.75\textheight]{visuals/fig05-time-to-grad.png}

\end{adjustwidth}








## Sankey diagram 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Numbers of students\newline   
Categorical: Starting major, destination major

\centering\includegraphics[height = 0.75\textheight]{visuals/fig005_ise_female.png}

\end{adjustwidth}
















## Diverging stacked bar chart 

\begin{adjustwidth}{-1.7em}{-1.7em}
Quantitative: Frequency of response, count totals\newline   
Categorical: Survey questions, level of confidence 

\centering\includegraphics[height = 0.7\textheight]{visuals/diverging-stacked-bar.png}


\scriptsize\raggedright 
\vertskip{0} 
Thomas Ryan, 2019 portfolio  
\end{adjustwidth}









## Multivariate bar chart 

\bigskip

\begin{columns}
\begin{column}{0.5\textwidth}

Quantitative
\begin{itemize}
\item number of people
\item percent became paralyzed
\end{itemize}

Categorical
\begin{itemize}
\item age group
\item vaccination status
\end{itemize}

\scriptsize
\vertskip{5} 
Few S (2014) Are mosaic plots worthwhile? Visual Business Intelligence Newsletter. Perceptual Edge https://tinyurl.com/y23majn2
\end{column}

\begin{column}{0.55\textwidth}
\includegraphics[width = 1\textwidth]{visuals/cm211-multivariate-bar.png}
\end{column}

\end{columns}

















## Linked micromaps 

```{r eval = FALSE}
library("magick")
im1 <- image_read("slides/visuals/d7-2.png")
im1 <- image_scale(im1, "1200")
im2 <- image_read("slides/visuals/d7-axis.png")
im2 <- image_scale(im2, "1200")
im  <- image_append(c(im1, im2), stack = TRUE)
image_write(im, path = "slides/visuals/d7-partial.png", format = "png")
```

\begin{adjustwidth}{-1.7em}{-1.7em}

\vertskip{0}
\includegraphics[height = 0.9\textheight]{visuals/d7-partial.png}

\scriptsize\raggedright 
\vertskip{-1} 
Nick Harrelson, 2019 portfolio    

\end{adjustwidth}









## Proportional symbol map 

\begin{adjustwidth}{-1.7em}{-1.7em}


\vertskip{-1} 
\includegraphics[height = 0.95\textheight, trim = 0mm 0mm 0mm 12mm, clip = TRUE]{visuals/d5.png}

\scriptsize\raggedright 
\vertskip{-2} 
Nick Harrelson, 2019 portfolio   

\end{adjustwidth}









## Dot density map 

Quantitative: Latitude, longitude \newline
Categorical: Race/ethnicity 

\begin{adjustwidth}{-1.7em}{-1.7em}

\vertskip{0} 
\includegraphics[height = 0.75\textheight]{visuals/cm211-dot-density-map-01.png}

\vertskip{-4} 
\raggedleft\includegraphics[height = 0.2\textheight]{visuals/cm211-dot-density-map-01-legend.png}

\scriptsize\raggedright 
\vertskip{-1} 
http://demographics.virginia.edu/DotMap/ 

\end{adjustwidth}
















## In summary, three precepts guide one's choice of graph type

\vertskip{0.5}
\begin{tabular}{@{} >{\centering\arraybackslash} m{42mm} m{66mm} @{}}

 \includegraphics[width = 42mm]{visuals/sd011-34.png} 
& \alertbf{Data structure} determines\newline which graph types are feasible\\

\includegraphics[width = 42mm, trim = 3mm 10mm 10mm 17mm, clip = TRUE]{visuals/sd011-34b.png}
& \alertbf{Graphical repertoire} determines\newline which graph types you explore\\

\includegraphics[width = 42mm, trim = 0mm 5mm 0mm 6mm, clip = TRUE]{visuals/align-visual-with-verbal.png}
& \alertbf{Design choices} align your visual argument\newline with your rhetorical goals 

\end{tabular}




 

 

 

 

 




```{r}
# pdf slide byproducts
unlink("slides/*.tex")
unlink("slides/*.aux")
unlink("slides/*.snm")
unlink("slides/*.toc")
unlink("slides/*.out")
unlink("slides/*.nav")
unlink("slides/*.log")
```

